---
title: "ACMG Evidence Codes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ACMG Evidence Codes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 14, fig.height = 8, dpi = 120)
```

Every `plot_variant_overlay()` figure carries a compact **ACMG strip**
directly under the variant header: one coloured chip per triggered
evidence code. This article explains what each chip means, exactly how
`compute_acmg_codes()` decides to draw it, and how to tune the one rule
that has a user-facing knob.

```{r demo-setup, message = FALSE}
library(msaVariant)

# Use a throwaway cache so this article never touches your real one.
Sys.setenv(MSAVARIANT_CACHE = tempfile("msaVariant_cache_"))

import_local_bundle(
  system.file("extdata", "DEMO1.rds", package = "msaVariant"),
  gene = "DEMO1"
)

demo_fasta <- system.file("extdata", "demo_aligned.fasta",
                          package = "msaVariant")
demo <- fetch_gene_data("DEMO1")
```

> **DEMO1 is synthetic.** The bundle used throughout this article is
> fabricated demonstration data — not real ClinVar, gnomAD,
> AlphaMissense, REVEL or CADD records. It was engineered so that one
> variant, `p.R21H`, fires all five codes at once, which makes it a
> convenient teaching example and a poor clinical one.

# The five codes

`msaVariant` implements a deliberately small, transparent subset of the
ACMG/AMP evidence framework. These are the codes that can be derived
*entirely from the tables already inside a gene bundle* — no extra data
source, no network call, no hidden model.

| Code | Strength | Plain-language meaning |
|------|----------|------------------------|
| **PS1** | Strong | This exact amino-acid change is already a known pathogenic variant. |
| **PM1** | Moderate | The residue sits inside an annotated functional domain. |
| **PM2** | Moderate | The change is absent from, or vanishingly rare in, the population. |
| **PM5** | Moderate | A *different* change at this same residue is already known to be pathogenic. |
| **PP3** | Supporting | The computational predictors agree the change is damaging. |

This is **not** a full ACMG classifier. It computes evidence codes; it
does not combine them into a Pathogenic/VUS/Benign verdict. That
judgement stays with the curator.

# How each code is derived

## PS1 — same change, already pathogenic

Fires when the ClinVar table contains a row whose `aa_change` matches
the query exactly **and** whose `significance` is `"Pathogenic"`.
`"Likely_pathogenic"` does *not* satisfy PS1 — only a full pathogenic
assertion does.

```{r ps1}
subset(demo$clinvar, aa_change == "R21H")
```

## PM1 — inside a functional domain

Fires when `variant_pos` falls within the `start`–`end` range of at
least one row of the `domains` table. PM1 is the only code that needs
nothing but a position, so it is evaluated even when you pass a
position-only label such as `"21"`.

```{r pm1}
demo$domains
```

Residue 21 sits inside the 5–35 "Demo functional domain", so PM1 fires.

## PM2 — absent or rare in gnomAD

Fires when the query `aa_change` is **absent** from the gnomAD table, or
present with `af_joint < 0.0001`. If a bundle carries no gnomAD table at
all, the variant is treated as absent and PM2 fires.

```{r pm2}
nrow(subset(demo$gnomad, aa_change == "R21H"))  # 0 rows -> absent -> PM2
```

## PM5 — different change, same residue, already pathogenic

Fires when ClinVar has a record at the **same position** with a
**different** alternate residue, rated `"Pathogenic"` *or*
`"Likely_pathogenic"`. Note the asymmetry with PS1: PM5 accepts the
likely-pathogenic tier, PS1 does not.

```{r pm5}
subset(demo$clinvar, pos == 21)
```

`R21A` (Pathogenic) and `R21C` (Likely_pathogenic) are both different
alternate residues at position 21, so PM5 fires for `R21H`.

## PP3 — the predictors agree

PP3 counts how many of three computational predictors call the
substitution damaging:

| Predictor | Passes when |
|-----------|-------------|
| AlphaMissense | `am_class == "likely_pathogenic"` |
| REVEL | `revel_score > 0.7` |
| CADD | `cadd_phred >= 25` |

```{r pp3-evidence}
subset(demo$alphamissense, aa_change == "R21H")[, c("aa_change", "am_score", "am_class")]
subset(demo$revel,         aa_change == "R21H")[, c("aa_change", "revel_score")]
subset(demo$cadd,          aa_change == "R21H")[, c("aa_change", "cadd_phred")]
```

All three pass for `R21H`. How many *need* to pass before PP3 fires is
the one rule you control — see below.

# Calling it directly

`compute_acmg_codes()` is a plain function. It takes a bundle, a
position and an `aa_change`, and returns the triggered codes in
canonical evidence-strength order:

```{r compute}
compute_acmg_codes(demo, variant_pos = 21, aa_change = "R21H")
```

The `p.` prefix is optional — `"p.R21H"` and `"R21H"` are equivalent.

# Tuning PP3 with `pp3_min_predictors`

`pp3_min_predictors` sets how many of the three predictors must pass:

* `1` — **any** predictor is enough (permissive)
* `2` — **majority**, the default
* `3` — **unanimous** (conservative)

Because `R21H` passes all three predictors, it is completely insensitive
to this setting — a useful property to be able to demonstrate:

```{r pp3-r21h}
for (k in 1:3) {
  cat("pp3_min_predictors =", k, "->",
      paste(compute_acmg_codes(demo, 21, "R21H", pp3_min_predictors = k),
            collapse = " "), "\n")
}
```

Variants with weaker computational support are where the knob earns its
keep. `I30A` passes AlphaMissense and CADD but fails REVEL (2 of 3):

```{r pp3-i30a}
for (k in 1:3) {
  cat("pp3_min_predictors =", k, "->",
      paste(compute_acmg_codes(demo, 30, "I30A", pp3_min_predictors = k),
            collapse = " "), "\n")
}
```

PP3 survives at `1` and `2`, and drops out at `3`. And `L27A`, which
only AlphaMissense flags (1 of 3), keeps PP3 only in permissive mode:

```{r pp3-l27a}
for (k in 1:3) {
  cat("pp3_min_predictors =", k, "->",
      paste(compute_acmg_codes(demo, 27, "L27A", pp3_min_predictors = k),
            collapse = " "), "\n")
}
```

The same argument is accepted by `plot_variant_overlay()` and is
forwarded straight through, so the chip strip in the figure always
matches what `compute_acmg_codes()` would return for the same settings.

# The strip in a figure

```{r figure, fig.alt = "DEMO1 p.R21H variant overlay figure with all five ACMG chips"}
plot_variant_overlay(
  gene          = "DEMO1",
  aligned_fasta = demo_fasta,
  variant_pos   = 21,
  variant_label = "p.R21H"
)
```

Two behaviours are worth knowing:

**Untriggered codes are hidden, not greyed out.** The strip shows only
the chips that fired. If nothing fires, the row renders a muted
`(no ACMG codes triggered)` placeholder rather than an empty band, so it
never reads as a rendering failure.

**Display is separate from computation.** `acmg_codes` controls which
chips are *drawn*; the full computation always runs underneath. To show
only the moderate-tier codes:

```{r subset-codes, fig.alt = "DEMO1 p.R21H figure showing only PM1, PM2 and PM5 chips"}
plot_variant_overlay(
  gene          = "DEMO1",
  aligned_fasta = demo_fasta,
  variant_pos   = 21,
  variant_label = "p.R21H",
  acmg_codes    = c("PM1", "PM2", "PM5")
)
```

To drop the strip entirely, set `show_acmg = FALSE` — see
`vignette("toggling-layers")`.

# Which codes need an alternate allele

PS1, PM2, PM5 and PP3 all compare a specific substitution, so they are
only evaluated when `aa_change` parses to a complete ref+pos+alt string
such as `"R21H"`. PM1 depends only on position and is always evaluated.

This matters when you plot a variant by position alone: you get PM1 and
nothing else, rather than a spuriously confident PM2 from a lookup that
could never have matched.

```{r pos-only}
compute_acmg_codes(demo, variant_pos = 21, aa_change = "21")
```

In `plot_variant_overlay()`, `aa_change` is taken from the explicit
`aa_change` argument if given, otherwise parsed from `variant_label`.
Passing `variant_label = "p.R21H"` is therefore enough to get the full
set; passing `variant_label = "Patient 4 variant"` is not, and you
should supply `aa_change = "R21H"` alongside it.

# See also

* `vignette("annotation-tracks")` — the tables these rules read from
* `vignette("toggling-layers")` — hiding the strip and individual tracks
* `?compute_acmg_codes` — the reference page
