---
title: "Bring Your Own Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bring Your Own Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 12, fig.height = 6, dpi = 120)
```

The shipped tracks cover the databases most variant-interpretation work
touches, but they will never cover a lab-internal assay, a paper's
supplementary table, or a predictor released last month.
`msaVariant` has two entry points for your own data:

* **`geom_track()`** — one ad-hoc strip on an alignment, no bundle
  needed
* **`import_local_bundle()`** — a full gene bundle, so a gene you
  assembled yourself works with `plot_variant_overlay()` and
  `compute_acmg_codes()` like any other

```{r demo-setup, message = FALSE}
library(msaVariant)

Sys.setenv(MSAVARIANT_CACHE = tempfile("msaVariant_cache_"))
demo_fasta <- system.file("extdata", "demo_aligned.fasta",
                          package = "msaVariant")
```

# `geom_track()` — an arbitrary per-residue strip

`geom_track()` returns ordinary ggplot2 layers, so it composes onto a
`ggmsa()` alignment plot. It needs a data frame with a `pos` column
(residue position in the reference) plus a value column you name.

## Continuous tracks

```{r continuous-data}
# Any per-residue score: a deep-mutational-scan readout, a conservation
# metric, a predictor of your own.
my_scores <- data.frame(
  pos   = 1:40,
  score = round(abs(sin(seq(0, 6, length.out = 40))), 3)
)
head(my_scores)
```

```{r continuous-layer, eval = FALSE}
library(ggmsa)

ggmsa(demo_fasta, char_width = 0.6, seq_name = TRUE) +
  geom_track(my_scores, msa = demo_fasta,
             value = "score",
             name  = "DMS fitness",
             type  = "continuous")
```

The chunk above is not evaluated here because `ggmsa` is a suggested
dependency, not a hard one. Install it with
`BiocManager::install("ggmsa")` to run it.

Useful arguments:

* `palette` — colours ramped low-to-high; defaults to a
  blue–yellow–red diverging ramp
* `value_range` — pin `c(min, max)` instead of fitting to the observed
  data, so several figures stay comparable
* `y_offset`, `track_height` — vertical placement in MSA-row units, for
  stacking more than one track
* `ref_name` — reference sequence name, matching the parent `ggmsa()`
  call

## Discrete tracks

Set `type = "discrete"` and give a named palette over the levels:

```{r discrete-layer, eval = FALSE}
my_calls <- data.frame(
  pos = c(6, 21, 30),
  sig = c("Benign", "Pathogenic", "VUS")
)

ggmsa(demo_fasta, char_width = 0.6, seq_name = TRUE) +
  geom_track(my_calls, msa = demo_fasta,
             value   = "sig",
             type    = "discrete",
             name    = "In-house panel",
             palette = c(Benign     = "#2166AC",
                         VUS        = "#9E9E9E",
                         Pathogenic = "#B2182B"))
```

Without a `palette`, discrete levels get an automatic hue palette;
values not found in a supplied palette fall back to grey.

## What `geom_track()` does with `pos`

Positions are mapped from reference residue numbering to alignment
columns with `map_variant_to_msa()`, exactly like the built-in tracks:

```{r mapping}
map_variant_to_msa(c(1, 21, 40), demo_fasta, ref_name = "DEMO1_HUMAN")
```

Positions that land on a gap in the reference map to `NA` and are
dropped. If *nothing* maps, you get a warning and `NULL` rather than a
silently empty strip — usually a sign that your table is in transcript
or genomic coordinates rather than protein residue numbering.

# `import_local_bundle()` — your own gene bundle

`geom_track()` adds a strip. A bundle gets you the whole machine: the
multi-panel figure, the legend dashboard and the ACMG evidence strip,
for a gene that has no Zenodo deposit.

```{r import}
path <- system.file("extdata", "DEMO1.rds", package = "msaVariant")
import_local_bundle(path, gene = "DEMO1")
```

The file is validated, then copied into the cache
(`cache_location()`), after which `fetch_gene_data()` finds it like
anything else — no network access:

```{r use-imported}
demo <- fetch_gene_data("DEMO1")
compute_acmg_codes(demo, 21, "R21H")
```

```{r plot-imported, fig.height = 8, fig.alt = "Full overlay figure built from a locally imported bundle"}
plot_variant_overlay(
  gene          = "DEMO1",
  aligned_fasta = demo_fasta,
  variant_pos   = 21,
  variant_label = "p.R21H"
)
```

`gene` is optional — omitted, the symbol is read from `bundle$meta$gene`
and used as the cache filename. Also available: `overwrite = FALSE` to
refuse replacing an existing bundle, and `verify_checksum` to compare
against a local `MANIFEST.tsv` when one is present (a mismatch warns but
still imports).

# The bundle schema

A bundle is a named list of seven data frames. The fastest way to see
the expected shape is to ask for an empty one:

```{r empty}
skeleton <- empty_gene_data(gene = "MYGENE", uniprot_id = "P00000",
                            protein_length = 393L)
str(skeleton, max.level = 1)
```

```{r empty-cols}
lapply(skeleton, names)
```

Build your tables into that skeleton and check as you go:

```{r validate}
skeleton$domains <- data.frame(
  start     = 5L,
  end       = 35L,
  name      = "My domain",
  accession = "PF00000",
  source    = "Pfam"
)

v <- validate_gene_data(skeleton, strict = FALSE)
v$valid
```

```{r validate-bad}
broken <- skeleton
broken$revel <- data.frame(pos = 1:3)   # missing revel_score
validate_gene_data(broken, strict = FALSE)$issues
```

`strict = TRUE` errors on the first problem instead of returning an
issue list — the right mode inside a build script.
`import_local_bundle()` runs `validate_gene_data(strict = FALSE)` itself
and refuses a non-conforming file, so a bad bundle fails at import
rather than halfway through drawing a figure.

Two things to get right, because validation cannot catch them:

**Numbering.** `pos` is a residue position in the canonical reference
protein — the same protein whose sequence is in your aligned FASTA. Not
transcript coordinates, not genomic.

**`aa_change` format.** `"R21H"` — reference residue, position,
alternate residue, no `p.` prefix. This string is matched *exactly* by
the PS1, PM2 and PP3 rules, so a format drift between your ClinVar table
and your CADD table means codes quietly fail to fire.

Once the tables validate, save and import:

```{r save-bundle, eval = FALSE}
saveRDS(skeleton, "MYGENE.rds")
import_local_bundle("MYGENE.rds")
```

# Partial bundles

Every table may be empty. A bundle with only `meta` and `domains` is
valid, and the figure re-flows to whatever is populated — an absent
table behaves exactly like `show_<track> = FALSE`
(see `vignette("toggling-layers")`).

The one asymmetry worth knowing: an empty gnomAD table means "never
observed in the population", so **PM2 fires for every variant**. If you
have no gnomAD data — as opposed to genuine absence from gnomAD — that
is a false PM2, and you should read the strip accordingly.

# Cache management

```{r cache}
cache_location()
cache_summary()
```

```{r clear, eval = FALSE}
clear_cache()          # remove everything
clear_cache("gnomad")  # remove one slice
```

Point `MSAVARIANT_CACHE` at a project directory to keep a study's
bundles beside the analysis, which is also how the examples in these
articles avoid touching your real cache:

```{r cache-env, eval = FALSE}
Sys.setenv(MSAVARIANT_CACHE = "~/projects/my_study/msaVariant_cache")
```

# See also

* `?geom_track`, `?import_local_bundle`, `?validate_gene_data`
* `vignette("annotation-tracks")` — what each built-in table holds
* `vignette("acmg-evidence-codes")` — the rules your tables will feed
