---
title: "Colour Schemes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Colour Schemes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 14, fig.height = 8, dpi = 120)
```

Journals have opinions about colour. `msaVariant` ships three presets
that cover the usual demands — a muted default, an explicit
colourblind-safe palette, and pure black-and-white — plus per-element
overrides for when a house style needs something specific.

```{r demo-setup, message = FALSE}
library(msaVariant)

Sys.setenv(MSAVARIANT_CACHE = tempfile("msaVariant_cache_"))
import_local_bundle(
  system.file("extdata", "DEMO1.rds", package = "msaVariant"),
  gene = "DEMO1"
)
demo_fasta <- system.file("extdata", "demo_aligned.fasta",
                          package = "msaVariant")
```

> DEMO1 is synthetic demonstration data, not real annotations. The same
> variant, `p.R21H`, is rendered in every scheme below so the only thing
> changing between figures is colour.

# The three presets

## `"journal"` — the default

Muted, professional, no neon. The ACMG chips run from dark red (strong
evidence) through orange to amber (supporting), which is both
semantically intuitive and monotonic in luminance.

```{r journal, fig.alt = "DEMO1 p.R21H rendered in the journal colour scheme"}
plot_variant_overlay(
  gene          = "DEMO1",
  aligned_fasta = demo_fasta,
  variant_pos   = 21,
  variant_label = "p.R21H",
  color_scheme  = "journal"
)
```

## `"colorblind"` — explicit Okabe–Ito

Draws from the Okabe–Ito qualitative palette, designed to stay
distinguishable under deuteranopia, protanopia and tritanopia. Reach for
this when a reviewer asks for a colourblind-safe figure by name, or when
the categorical ClinVar track is carrying a lot of the message.

```{r colorblind, fig.alt = "DEMO1 p.R21H rendered in the colorblind (Okabe-Ito) scheme"}
plot_variant_overlay(
  gene          = "DEMO1",
  aligned_fasta = demo_fasta,
  variant_pos   = 21,
  variant_label = "p.R21H",
  color_scheme  = "colorblind"
)
```

## `"grayscale"` — for black-and-white print

Every sequential ramp becomes a luminance ramp and the MSA tiles go flat
light grey, leaning on the bold residue letters rather than colour to
stay readable. Use this when a journal charges for colour figures, or to
check that a colour figure would still survive being photocopied.

```{r grayscale, fig.alt = "DEMO1 p.R21H rendered in the grayscale scheme"}
plot_variant_overlay(
  gene          = "DEMO1",
  aligned_fasta = demo_fasta,
  variant_pos   = 21,
  variant_label = "p.R21H",
  color_scheme  = "grayscale"
)
```

Comparing the three side by side is the fastest way to see the design
constraint the defaults are built around: the ACMG chips and every
continuous track stay ordered by darkness, so the grayscale version
loses hue but never loses the ranking.

# Per-element overrides

Every override is applied *on top of* the preset, so you change only
what you name and inherit the rest.

## ACMG chip colours

`acmg_colors` takes a named vector over `PS1`/`PM1`/`PM2`/`PM5`/`PP3`.
Partial vectors are fine — unnamed codes keep their preset colour.

```{r acmg-colors, fig.alt = "DEMO1 p.R21H with custom PS1 and PP3 chip colours"}
plot_variant_overlay(
  gene          = "DEMO1",
  aligned_fasta = demo_fasta,
  variant_pos   = 21,
  variant_label = "p.R21H",
  acmg_colors   = c(PS1 = "#2166AC", PP3 = "#92C5DE")
)
```

Chip label text is chosen automatically — black or white, whichever has
better contrast against the fill by WCAG relative luminance — so a dark
override will not leave you with unreadable text.

## Variant highlight

`variant_highlight_color` sets the fill of the column highlight that
runs through the header, the MSA and every track. The outline colour
stays with the scheme.

```{r variant-color, fig.alt = "DEMO1 p.R21H with a custom variant highlight colour"}
plot_variant_overlay(
  gene                    = "DEMO1",
  aligned_fasta           = demo_fasta,
  variant_pos             = 21,
  variant_label           = "p.R21H",
  variant_highlight_color = "#56B4E9"
)
```

## Track palettes

`track_palettes` is a named list accepting any of `clinvar`, `gnomad`,
`cadd`, `alphamissense` and `revel`. Continuous tracks take a vector of
colours ramped low-to-high; the categorical ClinVar track takes a vector
named by significance level.

```{r track-palettes, fig.alt = "DEMO1 p.R21H with custom CADD and ClinVar palettes"}
plot_variant_overlay(
  gene           = "DEMO1",
  aligned_fasta  = demo_fasta,
  variant_pos    = 21,
  variant_label  = "p.R21H",
  track_palettes = list(
    cadd    = c("#FFFFFF", "#9ECAE1", "#4292C6", "#08519C"),
    clinvar = c(Pathogenic        = "#000000",
                Likely_pathogenic = "#525252",
                VUS               = "#BDBDBD",
                Likely_benign     = "#DEEBF7",
                Benign            = "#9ECAE1",
                Conflicting       = "#7F7F7F",
                Other             = "#EEEEEE")
  )
)
```

## MSA residue colours

`msa_color_scheme` takes either a scheme name — `"journal"`,
`"colorblind"`, `"clustal"` or `"grayscale"` — or an explicit named
per-residue vector. This is independent of `color_scheme`, so you can
keep journal-muted tracks over a familiar Clustal-coloured alignment:

```{r msa-clustal, fig.alt = "DEMO1 p.R21H with Clustal MSA colouring and journal tracks"}
plot_variant_overlay(
  gene             = "DEMO1",
  aligned_fasta    = demo_fasta,
  variant_pos      = 21,
  variant_label    = "p.R21H",
  color_scheme     = "journal",
  msa_color_scheme = "clustal"
)
```

The three non-Clustal MSA palettes colour by amino-acid property group
(hydrophobic, positive, negative, polar, cysteine, glycine, proline,
aromatic) rather than per residue, which keeps a 20-colour legend from
swamping the figure.

# Saving for publication

`plot_variant_overlay()` returns a `patchwork` object, so
`ggplot2::ggsave()` handles output. Two settings cover most submission
requirements:

```{r save, eval = FALSE}
p <- plot_variant_overlay(
  gene          = "DEMO1",
  aligned_fasta = demo_fasta,
  variant_pos   = 21,
  variant_label = "p.R21H"
)

# Raster at publication resolution
ggplot2::ggsave("demo1_R21H.png", p,
                width = 15, height = 9, dpi = 300, bg = "white")

# Vector — preferred by most journals; text stays selectable
ggplot2::ggsave("demo1_R21H.pdf", p,
                width = 15, height = 9, device = grDevices::cairo_pdf)
```

Most journals want 300 DPI for raster figures and will accept — often
prefer — vector PDF. Because every panel is drawn with ordinary ggplot2
primitives, the PDF is genuine vector output: residue letters stay
crisp at any zoom, and a copy-editor can select the text.

If the target is a black-and-white page, render with
`color_scheme = "grayscale"` rather than letting the printer convert a
colour figure. The preset's ramps are luminance-monotonic by
construction, so evidence strength and score magnitude survive; an
automatic conversion of the `journal` palette can collapse distinct hues
onto near-identical greys.

# See also

* `vignette("toggling-layers")` — removing tracks to simplify a figure
* `?plot_variant_overlay` — every colour argument in one place
