---
title: "Toggling Layers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Toggling Layers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 14, fig.height = 8, dpi = 120)
```

A figure for a supervisor and a figure for a journal are rarely the same
figure. `plot_variant_overlay()` gives every annotation layer an on/off
switch, and the layout re-flows to whatever remains — no blank rows, no
orphaned axis, no legend for a track that is not there.

```{r demo-setup, message = FALSE}
library(msaVariant)

Sys.setenv(MSAVARIANT_CACHE = tempfile("msaVariant_cache_"))
import_local_bundle(
  system.file("extdata", "DEMO1.rds", package = "msaVariant"),
  gene = "DEMO1"
)
demo_fasta <- system.file("extdata", "demo_aligned.fasta",
                          package = "msaVariant")
```

> DEMO1 is synthetic demonstration data, not real annotations.

# The switches

| Argument | Layer |
|----------|-------|
| `show_acmg` | ACMG evidence-code chip strip |
| `show_domain` | Protein domain track |
| `show_clinvar` | ClinVar significance track |
| `show_gnomad` | gnomAD allele-frequency track |
| `show_cadd` | CADD PHRED track |
| `show_alphamissense` | AlphaMissense score track |
| `show_revel` | REVEL score track |

All default to `TRUE`. The variant header and the MSA panel have no
switch — they are the figure.

# Everything on

The default: header, ACMG strip, alignment, all six tracks, and a legend
dashboard.

```{r full, fig.alt = "Full DEMO1 p.R21H figure with all annotation layers"}
plot_variant_overlay(
  gene          = "DEMO1",
  aligned_fasta = demo_fasta,
  variant_pos   = 21,
  variant_label = "p.R21H"
)
```

# Everything off but the essentials

Alignment, variant, ACMG strip. Nothing else.

```{r minimal, fig.height = 6, fig.alt = "Minimal DEMO1 p.R21H figure: MSA, variant and ACMG strip only"}
plot_variant_overlay(
  gene               = "DEMO1",
  aligned_fasta      = demo_fasta,
  variant_pos        = 21,
  variant_label      = "p.R21H",
  show_domain        = FALSE,
  show_clinvar       = FALSE,
  show_gnomad        = FALSE,
  show_cadd          = FALSE,
  show_alphamissense = FALSE,
  show_revel         = FALSE
)
```

Compare the two. The minimal figure is not the full figure with holes in
it — the panel heights are recomputed, the residue-position axis has
migrated from the bottom track up onto the MSA panel, and the legend
dashboard has disappeared entirely because no legend-carrying track
remains. That is the whole point of the toggles: a subset stays
publication-ready.

# In-between: the in-silico predictors only

A common case — you want the computational evidence and the alignment,
without the clinical and population context.

```{r predictors, fig.height = 7, fig.alt = "DEMO1 p.R21H figure with only the in-silico predictor tracks"}
plot_variant_overlay(
  gene          = "DEMO1",
  aligned_fasta = demo_fasta,
  variant_pos   = 21,
  variant_label = "p.R21H",
  show_domain   = FALSE,
  show_clinvar  = FALSE,
  show_gnomad   = FALSE
)
```

Three tracks remain, three legends appear, and the axis sits on REVEL —
the lowest surviving track.

# Dropping the ACMG strip

`show_acmg = FALSE` removes the chip row. The codes are still computed
internally; they are simply not drawn.

```{r no-acmg, fig.alt = "DEMO1 p.R21H figure with the ACMG strip hidden"}
plot_variant_overlay(
  gene          = "DEMO1",
  aligned_fasta = demo_fasta,
  variant_pos   = 21,
  variant_label = "p.R21H",
  show_acmg     = FALSE
)
```

# Subsetting the codes instead

If you want the strip but not all five chips, keep `show_acmg = TRUE`
and pass `acmg_codes` — the display filter. Only codes that are both
requested *and* triggered are drawn, so asking for a code that did not
fire simply shows nothing rather than an empty chip.

```{r acmg-subset, fig.alt = "DEMO1 p.R21H figure showing only the PS1 and PM5 chips"}
plot_variant_overlay(
  gene          = "DEMO1",
  aligned_fasta = demo_fasta,
  variant_pos   = 21,
  variant_label = "p.R21H",
  acmg_codes    = c("PS1", "PM5")
)
```

The distinction is worth holding onto:

* `show_acmg` — does the strip exist at all?
* `acmg_codes` — which of the triggered chips get drawn?
* `pp3_min_predictors` — which codes trigger in the first place
  (see `vignette("acmg-evidence-codes")`)

Only the last one changes the underlying evidence call. The other two
are presentation.

# Missing data behaves like an off switch

A track is drawn only when its toggle is `TRUE` **and** the bundle
actually carries data for it. A bundle with no REVEL table produces the
same figure as `show_revel = FALSE` — the row is not reserved and then
left blank.

This means you can pass the same explicit toggle set across a whole
cohort of genes with uneven annotation coverage and every figure will
still come out tight.

# Windowing

Not a layer switch, but the other lever for figure density: `window`
sets the residue range on the x-axis, defaulting to ±20 residues around
the variant, clipped to the protein.

```{r window, fig.height = 7, fig.alt = "DEMO1 p.R21H figure restricted to residues 15-27"}
plot_variant_overlay(
  gene          = "DEMO1",
  aligned_fasta = demo_fasta,
  variant_pos   = 21,
  variant_label = "p.R21H",
  window        = c(15, 27),
  show_gnomad   = FALSE,
  show_cadd     = FALSE
)
```

A narrow window makes individual residue letters legible in a
two-column journal layout; a wide one shows more of the conservation
context. Both stay aligned to the same shared x coordinate system, so
every track lines up with the alignment column above it.

# See also

* `vignette("annotation-tracks")` — what each track shows
* `vignette("colour-schemes")` — restyling what is left after toggling
* `?plot_variant_overlay` — all arguments
